Product graphics

### 2010-04-25 product-graphics.tex
Thrashing paper into shape

\documentclass[letterpaper,oneside]{scrartcl}
\usepackage{fullpage}
\usepackage[utf8]{inputenc}
\usepackage[pdftex]{graphicx}
\usepackage{tikz}
\DeclareGraphicsExtensions{.png,.pdf}
\graphicspath{{plots/}}
\usepackage{hyperref}
\usepackage{url}
\usepackage[round,sectionbib]{natbib}
\bibliographystyle{abbrvnat}
\usepackage[small]{caption}
\usepackage[small]{titlesec}
\renewcommand\familydefault{bch}

@@

Area plots are the graphical equivalent of contingency tables, displaying how a total is broken down into components.

display categorical data with a polygon, typically a rectangle, for each combination of the factors of interest, with the area proportional to the number of observations in that combination.  Because the total area for a graphic is usually constrained, an area plot serves to partition up this total, displaying proportions rather than counts.

Figure \ref{fig:cat-examples} illustrates some of these plots.

\begin{figure}[htbp]
  \begin{center}
  %  \includegraphics[scale=1]{file}
  \end{center}
  \caption{Area plot examples.}
  \label{fig:cat-examples}
\end{figure}

Hierarchical partitioning of space.

\section{Factorising}

Basic idea is factorising table of probabilities:

Some notation: let capital letters refer to the sample space, and lower case particular values.

It can be useful to display partial decompositions, i.e. just $f(x, y | z)$, which is still a 3 variable pmf, and is defined as

\[ f(x, y | z) = \frac{f(x, y, z)}{\sum_{x = a, y = b} f(a, b, z)} \]

so that each category of z sums to one, and for any fixed value of z, $f(x, y | z)$ is a valid pmf.

\[ \forall b \sum_a f(a | b) = 1 \]

Because of their hierarchical nature, an area plot of $f(x, y, z)$ also shows $f(y, z)$ and $f(z)$.

Figure~\ref{fig:fact-simple} illustrates these ideas with the distribution of happiness and sex. We start with a plot where area broken down by happiness. We can then continue to factor this plot to also display sex - the area of each rectangle is proportional to the number of people who are happy with that sex, and the construction allows us to see the conditional distribution of sex within happy. In this case the breaks line up, suggesting that there is no difference in sex, given happiness.  This could be followed up by a formula hypothesis test.

Removing the marginal effect of happiness, as in the column on the far right, makes each of the top-level categories the same size, and makes it easier to compare the conditional distributions.  It's not that important in this case, but in the bar chart below, it allows us to see the conditional distribution, and in the case when there are many levels with very different probabilities it can make the plot much easier to read.

\begin{figure}[htbp]
  \centering
  \includegraphics[width=0.33\linewidth]{fact-happy}%
  \includegraphics[width=0.33\linewidth]{fact-happy-sex}%
  \includegraphics[width=0.33\linewidth]{fact-happy|sex}

  \includegraphics[width=0.33\linewidth]{fact-happy-2}%
  \includegraphics[width=0.33\linewidth]{fact-happy-sex-2}%
  \includegraphics[width=0.33\linewidth]{fact-happy|sex-2}

% Hierarchical coordinate systems 
%
% What are the common components of all these plots, and how can be describe
% them in a succinct way that also allows us to describe what tasks each
% display is good for?  We propose that there is one key feature that makes
% these plots special.  That is there their coordinate system, which is
% hierarchical.  
%
% In most coordinate systems that we are used to, such as Cartesian or polar
% coordinates, are symmetric in the sense that you can locate a point by
% starting from either dimension.  For example, in Cartesian coordinates you
% can identify either the x- or y-coordinate first, and the other next.  This
% also implies a distance symmetry when projected onto 2D Euclidean geometry.
% When you vary one coordinate by a small amount (holding all others constant)
% points in the projective geometry will be close.
%
% Neither of these properties hold for hierarchical coordinate systems. 
%
% We will discuss hierachical coordinate systems in general, and then present
% a specific hierachical coordinate system which gives rise to the area plots
% described previously.

\section{Data}
\label{sec:data}

Two ways of representing data:

\begin{itemize}
  \item Multi-dimensional contingency table
  \item Data frame, with column of counts
\end{itemize}

The two are equivalent, but it's usually computational easier to work with the data frame version, because variables are explicit, and mathematically easier to work with the multidimensional array because multiplication is nicely defined.  Also deals better with sparsity and is common input to modelling algorithms.

Define the conditioning operator which makes the distribution uniform.

In this paper, we'll use a small sample from the general social survey ({\sc gss}) focussing on happiness. 51\,020 observations from cross-section survey given yearly from 1972 to 2006.

\begin{itemize}
  \item Happy.  Discrete: ``very happy'', ``pretty happy'', ``not too happy''
  \item Age (in years), 18--89.
  \item Sex: ``female'', ``male''
  \item Degree: ``lt high school'', ``high school'', ``junior college'', ``bachelor'', ``gradudate''
  \item Relative financial status {\sf finrela}: ``far above average'', ``above average'', ``average'', ``below average'', ``far below average''
  \item Health: ``excellent'', ``good'', ``fair'', ``poor''
  \item Year: 1972--2006.
  \item Probability weight ({\sf wtsall}): 0.43--6.42.
\end{itemize}

\section{Display}
\label{sec:display}

Once we have made the decision to map area to proportion, we need to impose some constraints on the set of possible tilings of the plane. In general, we could also think of partitions of higher-dimensional spaces (e.g. 3d or 4d), but we will need to project these down to two for viewing anyway, so there is little disadvantage to working in 2d.

Some constraints are implied by our perceptual capabilities and others by geometry. On top of the constraint that area must be proportional to value, we add two additional constraints:

  \item Partitions must be rectangular. It is easier to compare the areas of simple shape (i.e. rectangle vs polygon), and even better when one the length of one border is fixed, so we only need compare the length of the other \citep{cleveland:1984}. Rectangular partitions are also recursive in the sense that we can always tile a rectangle with smaller rectangles. This is not true for most shapes.

  \item Partitions must be disjoint. To be able to see the complete area of each rectangle they must be non-overlapping, and each split must be nested within the shapes of the previous level. Note that we do not impose the constraint that the tiling be space-filling. Space filling methods use will ensure that areas are larger, but make it harder to compare magnitudes because scales no longer align.

@@

Note, however, that each of these constraints can be profitably relaxed. See Section~\ref{sec:relax} for details.

Let $a(i)$ give the area assigned to category $i$, which has proportion $f(i)$. Since we are using rectangles, $a(i) = k \dot a(i) = k \dot w(i) \times h(i)$ where $w$ and $h$ represent width and height respectively. Certain partitions also satisfy the constraint that $h(i) = \sqrt{k} f(i)^x$ and $w(i) = \sqrt{k} f(i)^y$ where $x + y = 1$. This is useful because it is only necessary to inspect one border to compare magnitudes.

The following two sections describe the partitions currently used to produce area graphics. Section~\ref{sub:part-1d} discusses partitions for 1d pmfs: bars, spines and tiles, Section~\ref{sub:part-2d} discusses one partition for 2d pmfs and Section~\ref{sub:part-nd} describes how the 1d and 2d primitives can be recursively combined to generate a partitioning for any higher dimension pmf and discusses how existing area graphics fall into this framework.

@@

  \item {\bf bar}s: height is proportional to value, width equally divides space. Bars allow comparison between absolute numbers, but are not space filling.  They occupy sum(values / max(values)) of the total area.  Bars can be arranged horizontally (``hbar'') or vertically (``vbar'').  Not space filling.

  \item {\bf spine}s: width is proportional to value, height equally divides space. Splines are space filling, but make it more difficult to compare absolute counts.  Splines can be arranged horizontally (``hspine''),  vertically (``vspine''), or can automatically pick their orientation (``spine'').  Space filling.

  \item {\bf tile}s: tile the plane with rectangles, trying to keep the aspect ratio of each rectangle close to 1. Comparing the areas is the most difficult perceptually, and are most useful when there are a large number of categories.  The is partitioning used by the squarified treemap \citep{bruls:1999}.  Not space filling.

@@

@@

We are currently aware of one 2d primitive, the {\bf fluct}: where width and height are proportional to the square root of the proportion. Flucts allow comparisons of relative magnitudes numbers in two directions. Each rectangle is arranged on a regular grid formed by the levels of the two variables, and thus the fluct is not space filling. A fluct displays a joint 2d distribution, unlike the combination of any two 1d partitions which display a conditional distribution.

@@

\begin{figure}[htbp]
  \centering
    \includegraphics[width=0.5\linewidth]{part-fluct}%
    \includegraphics[width=0.5\linewidth]{part-fluct-cond}
  \caption{(Left) Distribution of age (in decades) and marital status.  All combinations of the two variables are present. (Right) Distribution of health status and year of survey. Health status was not recorded for three years.}
  \label{fig:fluct}
\end{figure}

\subsection{Combinations}
\label{sub:part-nd}

To visualise higher-dimensional pmfs, we recursively partition using the previously described primitives. That is, to display $f(x, y, z)$ we first create a partition based on $f(z)$ then partition each of those pieces based on $f(y | z)$ and again by $f(x | y, z)$. Figure \ref{fig:recursive} illustrates this process with a graphic that explores whether, conditional on marital status, there is a difference in happiness between men and women.

% Heike: may be this would be a good spot for some of your Kronecker
% product theory?

@@

@@

Conditioning is also an important tool, because it allows one to remove relationships that are well-known or uninteresting. Figure~\ref{fig:part-cond} uses a fluct and a vspine to explore the relationship between happiness, health and financial status. The left plot displays raw proportions, showing that most people are in good health and average financial standing. However, it is difficult to see how happiness varies within these conditions because the perceptual task is comparing areas. Conditioning on financial status and health produces the plot on the right (known as an equal bin size plot) and makes it easier to see the conditional distribution of happiness, because the perceptual task is simpler: comparing length on a common non-aligned scales.  Here we can see people tend to be happier the healthier and richer they are.

@@

\section{Relaxing constraints}
\label{sec:relax}

There are a number of cases in which it is useful to relax the constraints that we defined above.  These are described in sequence below.

@@

On the other end of the spectrum, it can be useful to constrain put a cap on the largest possible values \citetext{Antony Unwin, priv. comm.}. By controlling this cap it is possible to zoom in on the smallest values.

Other non-linear transformations may also be useful.  For example, we could display the square root of the weights, to stabilise the variance of the areas.  If values span multiple orders of magnitude it may be useful to map log value to area.

@@

\section{Variations}
\label{sec:variations}

On top of this basic framework, there are a few variations designed for dealing with special types of data: missing data, weighted data, and continuous data.  These variations are described below.

This gives rise to the histogram and spinogram, analogues of the bar chart and spine chart respectively. A long standing tradition is that no gaps are displayed between adjacent rectangles when used for originally continuous data. Two examples are shown in Figure \ref{fig:cont-examples}.

\begin{figure}[htbp]
  \begin{center}
  \end{center}
  \caption{Area plots of originally continuous data}
  \label{fig:cont-examples}
\end{figure}

Give a few examples of non-traditional continuous-categorical plots.  Fluctuation diagram for exploring joint 2d distribution.  Mosaic for conditional.  Age and year distribution.

Cat + continuous.

An alternative approach is use density estimates (or conditional density estimates) and plot with continuous area plots.  As in: Hofmann, H., Theus, M. (2005), Interactive graphics for visualizing conditional distributions, Unpublished Manuscript.

% \subsection{Nested data}
% \label{sub:nested_data}
% 
% Use space-filling!
% 
% We distinguish nested data from crossed data by the number of interactions with missing values.  Completely crossed data contains every combination of all variables.  Nested data contains 
% 
% 
% Aka sparse vs. dense.
% 
% It is not always possible to determine nested data from inspection of the data alone, but may need knowledge of the data collection.  For example, if a survey of teachers within schools was performed, and each teacher was identified by a identifier unique within a school, it might look like the data was only missing a few combinations, when in fact it is meaningless.  
% 
% 
% 
% Or when you have a hierarchy.
% 
% Example dataset - Doug Bates teachers/schools?
% 
% Possibilities: create new variable which is interaction of nested vars, or drop missings/zeros.  Treemaps deal almost exclusively with nested data.

\section{Labelling}
\label{sec:legends}

Due to the hierarchical nature of these plots, it's difficult to label them in an informative way. We take a pragmatic approach, and label a single horizontal and single vertical split, where possible, using colour to identify a third variable, if necessary.

Understanding the composition of these plots is most easy when:

  \item Exploration: You have interactive control over the creation of the plot, and can interactively query regions of interest

  \item Communication: The plot is built up step-by-step, and shows the minimum number of splits.  It may be easier to understand a single complex plot when factored into two or more simpler plots.

To create the labels in used in this paper, we inspect each level of the factorisation looking for the first level in which there are rows or columns.  We do this rather inspecting the partitions that make up the plot, because there are many possible combinations that create columns: hsplines, hbars, vsplines, vbars for uniform data, flucts, ... and then can occur at any level in the hierarchy depending on what's come before.

Apart from textual labels on the x-axis, other work has labelled individual cells with attributes of the data:

\begin{itemize}
  \item colour (map of the market)
  \item texture, such as in sieve plots \citep{friendly:2000}
  \item photographs \citep{bederson:2001}
  \item text (tables)
  \item embedded plots (time series in lab escape)
\end{itemize}

Or attributes of the hierarchy:

\begin{itemize}
  \item Spacing / borders
  \item Shading
  \item Cascading
  \item Labelling
\end{itemize}

\section{Conclusions} % (fold)
\label{sec:conclusions}

% section conclusions (end)

>>>>

\documentclass[letterpaper,oneside]{scrartcl}
\usepackage{fullpage}
\usepackage[utf8]{inputenc}
\usepackage[pdftex]{graphicx}
\usepackage{tikz}
\DeclareGraphicsExtensions{.png,.pdf}
\graphicspath{{plots/}}
\usepackage{hyperref}
\usepackage{booktabs}
\usepackage{url}
\usepackage[round,sectionbib]{natbib}
\bibliographystyle{abbrvnat}
\usepackage[small]{caption}
\usepackage[small]{titlesec}
\renewcommand\familydefault{bch}

@@

Area plots, like bar charts and mosaic plots, are the graphical equivalent of contingency tables, displaying how a total is broken down.

Display of discrete distribution, pmf, contingency table.

This paper presents a unifying framework for area plots, showing how many existing plots are interrelated, and by drawing an

The paper proceeds as follows. Section~\ref{sec:primitives} defines the simple one- and two-d primitives used for all more complicated graphics. Section~\ref{sec:combination} provides a mathematical framework that shows how these simple primitives can be combined to display higher-d tables. Section~\ref{sec:variations} shows how these plots can be adapted to deal with missing values, weights and continuous data.

To illustrate these ideas, the same data for all examples. It's a small sample
of variables related to happiness from the general social survey ({\sc gss}).
The {\sc gss} is a yearly cross-sectional survey of Americans. Combining
surveys from 1972 to 2006 gives 51\,020 observations. Of the over 5\,000
variables, we select eight related to happiness, as described in
Table~\ref{tbl:happy}. The data is included in the supplemental materials.

\begin{table}
  \begin{center}
  \begin{tabular}{llp{8cm}}
    \toprule
    Variable name & Description & Values \\
    \midrule
    {\sf age} & age in years & 18--89 \\
    {\sf degree} & highest education & lt high school, high school, junior college, bachelor, graduate \\
    {\sf finrela} & relative financial status & far above average, above average, average, below average, far below average \\
    {\sf happy} & happiness & \key{very-happy} very happy, \key{pretty-happy} pretty happy, \key{not-too-happy} not too happy \\
    {\sf health} & health & excellent, good, fair, poor \\
    {\sf sex} & sex & \key{female} female, \key{male} male\\
    {\sf wtsall} & probability weight & 0.43--6.42 \\
    {\sf year} & year of survey & 1972--2006 \\
    \bottomrule

  \end{tabular}
  \end{center}
  \caption{Description of sample data.  Common plot colours are shown next to respective levels.}
  \label{tbl:happy}
\end{table}

\section{Graphical primitives}
\label{sec:primitives}

Tilings of the plane. In general, we could also think of partitions of higher-dimensional spaces (e.g. 3d or 4d), but we will need to project these down to two for viewing anyway, so there is little disadvantage to working directly in 2d. To derive the graphical primitives we start with three constraints:

\begin{itemize}
  \item Area must be {\bf proportional} to value. Because the total area for a graphic is usually constrained, an area plot serves to partition up this total, displaying proportions rather than counts.

  \item Partitions must be {\bf rectangular}. It is easier to compare the areas of simple shape (i.e. rectangle vs polygon), and even better when one the length of one border is fixed, so we only need compare the length of the other \citep{cleveland:1984}. Rectangular partitions are also recursive in the sense that we can always tile a rectangle with smaller rectangles. This is not true for most shapes.

  \item Partitions must be {\bf disjoint}. To be able to see the complete area of each rectangle they must be non-overlapping, and each split must be nested within the shapes of the previous level. Note that we do not impose the constraint that the tiling be space-filling. Space filling methods use will ensure that areas are larger, but make it harder to compare magnitudes because scales no longer align.

@@

\noindent Each of these three constraints can be profitably relaxed, as detailed in Section~\ref{sec:variations}.

The following sections describe the partitions generated by these constraints. Section~\ref{sub:part-1d} discusses 1d partitions, bars, spines and tiles, and  Section~\ref{sub:part-2d} discusses the 2d partition, the fluct.

% Let $a(i)$ give the area assigned to category $i$, which has proportion $f(i)$. Since we are using rectangles, $a(i) = k \dot a(i) = k \dot w(i) \times h(i)$ where $w$ and $h$ represent width and height respectively. Certain partitions also satisfy the constraint that $h(i) = \sqrt{k} f(i)^x$ and $w(i) = \sqrt{k} f(i)^y$ where $x + y = 1$. This is useful because it is only necessary to inspect one border to compare magnitudes.

@@

  \item {\bf bar}s: height is proportional to value, width equally divides space. Bars allow comparison between absolute numbers, but are not space filling. Bars are not space filling, occupying $\sum(p_i / \max(p_i))$ of the total area. Bars can be arranged horizontally (``hbar'') or vertically (``vbar'').

  \item {\bf spine}s: width is proportional to value, height is fixed. Splines are space filling. Splines can be arranged horizontally (``hspine''), vertically (``vspine''), or can automatically pick their orientation (``spine'').

  \item {\bf tile}s: tile the plane with rectangles, trying to keep the aspect ratio of each rectangle close to 1. There are no restrictions on height or width, just the overall area. This is partitioning used by the squarified treemap \citep{bruls:1999}.

@@

Each of these three displays has different strengths and weaknesses.  The perceptual task associated with bar charts is the simplest, but they do not occupy all the area. Spines and tiles are harder to read, but occupy the complete space and work better recursively \citep{heer:2010}.

@@

We are currently aware of one 2d primitive, the {\bf fluct}, which has both height and width proportional to the square root of the proportion. Flucts allow comparisons of relative magnitudes in two directions. Each rectangle is arranged on a regular grid formed by the levels of the two variables, and so is not space filling.

@@

\begin{figure}[htbp]
  \centering
    \includegraphics[width=0.5\linewidth]{part-fluct}%
    \includegraphics[width=0.5\linewidth]{part-fluct-cond}
  \caption{(Left) Distribution of age (in decades) and marital status. (Right) Equal bin-size plot showing health status and survey year. Health status was not recorded for three years.}
  \label{fig:fluct}
\end{figure}

\section{Factorising area plots}
\label{sec:combination}

To construct plots for discrete pdfs with more than two dimensions, we need a mechanism to factorise them into low dimensional products. This factorisation has three roles:

% It can be useful to display partial decompositions, i.e. just $f(x, y | z)$, which is still a 3 variable pmf, and is defined as
%
% \[ f(x, y | z) = \frac{f(x, y, z)}{f(z)} \]
%
% so that each category of z sums to one, and for any fixed value of z, $f(x, y | z)$ is a valid pmf.
%
% \[ \forall b \sum_a f(a | b) = 1 \]
%
% Because of their hierarchical nature, an area plot of $f(x, y, z)$ also shows $f(y, z)$ and $f(z)$.

Figure~\ref{fig:fact-simple} uses the distribution of sex and happiness to illustrates these ideas. From left to right, we have plots of \verb|~ happy|, \verb!~ happy | sex! and their product \verb|~ happy + sex|.  Observe that the heights and widths of the rectangles are multiplied in the same way as the components of the pmf.

\begin{figure}[htbp]
  \centering
  \includegraphics[width=0.3\linewidth]{fact-happy} $\times$ %
  \includegraphics[width=0.3\linewidth]{fact-happy|sex} = %
  \includegraphics[width=0.3\linewidth]{fact-happy-sex}

  \includegraphics[width=0.3\linewidth]{fact-happy-2} $\times$ %
  \includegraphics[width=0.3\linewidth]{fact-happy|sex-2} = %
  \includegraphics[width=0.3\linewidth]{fact-happy-sex-2}

% To visualise higher-dimensional pmfs, we recursively partition using the previously described primitives. That is, to display $f(x, y, z)$ we first create a partition based on $f(z)$ then partition each of those pieces based on $f(y | z)$ and again by $f(x | y, z)$.

Figure \ref{fig:recursive} illustrates this process with a graphic that explores whether, conditional on marital status, there is a difference in happiness between men and women.

@@

Conditioning is also an important tool, because it allows one to remove relationships that are well-known or uninteresting. Figure~\ref{fig:part-cond} uses a fluct and a vspine to explore the relationship between happiness, health and financial status. The left plot displays raw proportions, showing that most people are in good health and average financial standing. However, it is difficult to see how happiness varies within these conditions because the perceptual task is comparing areas. Conditioning on financial status and health produces the plot on the right (an equal bin size plot) and makes it easier to see the conditional distribution of happiness, because the perceptual task is simpler: comparing length on a common non-aligned scales.  Here we can see people tend to be happier the healthier and richer they are.

\subsection{Existing plot types}

@@

@@

\section{Modelling}
\label{sec:models}

\section{Variations and extensions}
\label{sec:variations}

On top of this basic framework, it is useful to consider a few variations and extensions.  Three are based on variations of the underlying data (missing data, weighted data, and continuous data), and three are based on relaxing the display constraints (area not proportional to weight, non-disjoint partitions, non-rectangular partitions).

This gives rise to the histogram and spinogram, continuous analogues of the bar chart and spine chart. A long standing tradition is that no gaps are displayed between adjacent rectangles when used for continuous data. Two examples are shown in Figure \ref{fig:cont-examples}.

\begin{figure}[htbp]
  \centering
  \caption{Area plots of originally continuous data}
  \label{fig:cont-examples}
\end{figure}

Example exploring joint and conditional distributions of age and year.

% An alternative approach is use density estimates (or conditional density estimates) and plot with continuous area plots.  As in: Hofmann, H., Theus, M. (2005), Interactive graphics for visualizing conditional distributions, Unpublished Manuscript.

@@

On the other end of the spectrum, it can be useful to constrain put a cap on the largest possible values \citetext{Antony Unwin, priv. comm.}. By controlling this cap it is possible to zoom in on the smallest values. - called censored zooming by Antony and Martin.

Other non-linear transformations may also be useful. For example, we could display the square root of the weights, to stabilise the variance of the areas. Tukey applied this technique to histograms to create rootograms.

If values span multiple orders of magnitude it may be useful to map log value to area.

@@

\section{Conclusions}
